Physcraper: A Python package for continually updated gene trees

### 2020-05-07 physcraper_ms.Rmd
more examples

---
# Example from https://joss.readthedocs.io/en/latest/submitting.html
title: 'Physcraper: A Python package for continually updated gene trees'
tags:
  - Python
  - phylogenetics
  - update
  - species tree
  - Open Tree
authors:
  - name: Luna L. Sanchez Reyes
    orcid: 0000-0000-0000-0000
    affiliation: 1 # (Multiple affiliations must be quoted, like this: "1, 2")
  - name: Emily Jane McTavish
    orcid: 0000-0000-0000-0000
    affiliation: 1
affiliations:
 - name: University of California, Merced
   index: 1
 - name: Institution 2
   index: 2
citation_author: Sanchez-Reyes et. al.
date: 24 April 2020
year: 2020
bibliography: paper.bib
output: rticles::joss_article
csl: apa.csl
journal: JOSS
---

@@

1. We hope the physcraper workflow encourages researchers on their data sharing practices.

@@

# Physcraper

# Examples

Updating trees from @schoch2009ascomycota

1. Get study id from Open Tree of Life. There are some options to do this:
  - You can go to the Open Tree of Life web browser, or
  - you can do it using R:
    - By getting the treebase ID of the study (which is not fully exposed on the Treebase website home page of the study, so you have to look it up manually)
```{r}
rotl::studies_find_studies(property = "treebaseId", value = "S2137")
```

1. Download the alignment from Treebase
If you are on the Treebase home page of the study, you can navigate to the matrix tab, and manually download the alignments that were used to reconstruct the trees reported on the study that were also uploaded to Treebase and to the Open Tree of Life repository.
  To make this task easier, you can use a command to download everything into your working folder:
```
auto_scrape.py -s pg_238 -t tree109 -o ../physcraper_example/treebase_alns
```
In this example, all alignments posted on Treebase were used to reconstruct both trees.

@@

```{bash}
opentree_scrape.py -s pg_238 -t tree109 -a treebase_alns/pg_238-LSU-M3956.nex -as "nexus" -o tmp
```

@@

Tools that do similar things:
pyPhlawd [@smith2019pyphlawd]
SUPERSMART [@antonelli]

>>>>

---
# Example from https://joss.readthedocs.io/en/latest/submitting.html
title: 'Physcraper: A Python package for continually updated gene trees'
tags:
  - Python
  - phylogenetics
  - update
  - species tree
  - Open Tree
authors:
  - name: Luna L. Sanchez Reyes
    orcid: 0000-0000-0000-0000
    affiliation: 1 # (Multiple affiliations must be quoted, like this: "1, 2")
  - name: Emily Jane McTavish
    orcid: 0000-0000-0000-0000
    affiliation: 1
affiliations:
 - name: University of California, Merced
   index: 1
 - name: Institution 2
   index: 2
citation_author: Sanchez-Reyes et. al.
date: 24 April 2020
year: 2020
bibliography: paper.bib
output: rticles::joss_article
csl: apa.csl
journal: JOSS
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  eval = TRUE
)
```

@@

1. We hope the physcraper workflow demonstrates the benefits of open results in phylogenetics and encourages researchers to strive for better data sharing practices.

@@

# Physcraper overview

# Use case/ example

Imagine you are starting to work on a new biological group X. You have not much of an idea about
its phylogenetic relationships, you are a newly established researcher, and the group
is not anything any of your collaborators have worked on before.
A good idea is to start an intensive litt review on the phylogenetics of the group.
Rapidly, you find out there are 5 different phylogeneis, that used different markers, and that
the papers, published at different times, do not discuss which phylogeny is
the one accepted by the expert community on X. You might need
to go to the annual conference of X, and even then, you might only find
different and contrasting opinions.
Somewhere along these months or even years doing this task, you looked into the
the OToL database. You found in there some or all the published trees of X, along
with a tree that has been deemed the best tree by curators and ideally experts on X?

## Ascomycota Example

Let's be more specific now about our X group and say it is the Ascomycota.
The best tree currently available in OToL was published by @schoch2009ascomycota.
The first step, is to get the Open Tree of Life study id. There are some options to do this:
- You can go to the Open Tree of Life website and browse until you find it, or
- you can get the study id using R tools:
  - By using the treebase ID of the study (which is not fully exposed on the
    Treebase website home page of the study, so you have to really look it up manually):
```{r, treebase, eval = TRUE}
rotl::studies_find_studies(property = "treebaseId", value = "S2137")
```
  - By using the name of the focal clade of study (but this behaved very differently):
```{r, focal, eval = FALSE}
rotl::studies_find_studies(property="ot:focalCladeOTTTaxonName", value="Ascomycota")
```

Once we have the study id, we can gather the trees published on that study:
```{r, treeids, eval = TRUE}
rotl::get_tree_ids(rotl::get_study_meta("pg_238"))
rotl::candidate_for_synth(rotl::get_study_meta("pg_238"))
my_trees <- rotl::get_study("pg_238")
```
Both trees from this study have `r ape::Ntip(my_trees[[1]])` tips.

Let's check what one of the trees looks like:

```{r fungi-phylo, eval = TRUE, fig.retina = 2, fig.height = 4}
ape::plot.phylo(my_trees[[1]], type = "fan", cex = 0.1)
```

1. Download the alignment from Treebase
If you are on the Treebase home page of the study, you can navigate to the matrix tab, and manually download the alignments that were used to reconstruct the trees reported on the study that were also uploaded to Treebase and to the Open Tree of Life repository.
To make this task easier, you can use a command to download everything into your working folder:
```
physcraper_run.py -s pg_238 -t tree109 -o ../physcraper_example/pg_238
```
In this example, all alignments posted on Treebase were used to reconstruct both trees.

@@

```
physcraper_run.py -s pg_238 -t tree109 -a treebase_alns/pg_238tree109.aln -as "nexus" -o pg_238
```

## Testudines example

Phylogeny of the Testudines 6 tips from @crawford2012more
There is just one tree in OToL.
There is just one alignment on [treebase](https://treebase.org/treebase-web/search/study/matrices.html?id=12742) with all the 1 145 loci.

```
physcraper_run.py -s pg_2573 -t tree5959 -tb -db ~/branchinecta/local_blast_db/ -o pg_2573
```

@@

Tools that do similar things:
pyPhlawd [@smith2019pyphlawd]
SUPERSMART [@antonelli2017toward]
